On-the-fly Algorithms for GKAT Equivalences

### 2025-03-18 rebuttal.md

# Overall Response

We would like to express our sincere gratitude for the constructive feedback received from the reviewer. We will carefully address the points raised and implement the necessary changes to our paper.

First, we address the efficiency of our lazy normalization approach. 
The classical algorithm guarantees a full traversal of the coalgebra during liveness checking prior to bisimulation checking, our algorithm invokes bisimulation and only run traverses the reachable states of the discrepancy state.
Therefore, our lazy liveness checking consistently traverses fewer states, even for positive results. 
Additionally, lazy liveness checking enables on-the-fly coalgebra generation using methods like derivatives, which the original algorithm does not support.

Second, we address the theoretical contribution of our work. 
While the result may seem practically intuitive, we maintain that our proof is not trivial. 
We are able to provide straightforward and concise proofs because we introduced a middle ground between coalgebraic theory and practical algorithm, namely greedy bisimulation. 
Without this framework, verifying Theorem 22 and Theorem 23 would be more challenging, especially given that neither the input coalgebra nor their greedy bisimulation are assumed to be finite.

Finally, we emphasize that our contribution extends beyond bisimulation of machines (or coalgebras).
In Section V, we present two methods for generating coalgebras from expressions and prove their correctness, finiteness, and correspondence.
The finiteness result, in particular, arises naturally from the correspondence result. 
The simplicity of this result is noted by one of the reviewers.

Regarding the practical aspects of our work, specifically the experiments, we are committed to revising the tests to better represent real-world scenarios before the deadline. 
However, we also argue that our theoretical results have already demonstrated improvements over existing methods. 
Plus, trivial test cases do not provide any performance benefit for our algorithm based on Thompson's construction, and dead test cases also do not favor our algorithm against other tools. 
Although we will improve the balance and coverage of our test cases, we believe the current test cases are sufficient in demonstrating the efficiency of our implementation.

# Reviewer Questions

### Reviewer 1

> Where are the benchmarks from?

The benchmark is generated by qcheck library of Ocaml, using the module QCheck2. 
The generation is based on algebraic rules of GKAT (except the fixpoint rules) with reflexivity and symmetry rule.
We will include the generation script in the final submission, and provide documentation on test generation

> Unnecessary but harmful generalization

We will specify in the background section that we assume the function `F` to be simple polynomial functors.

### Reviewer 2

Our approach to symbolic transducers diverges from previous works in several key aspects:
- Our on-the-fly method is tailored to handle normalization, a unique challenge in achieving finite trace equivalence for GKAT. 
- The signature of GKAT differs from traditional symbolic transducers. Unlike typical accepting or rejecting states, a GKAT state can accept or reject based on different input atoms. This results in a distinct semantics where a trace's end is marked by an atom, representing the output program state.

While symbolic bisimulation is pivotal for the efficiency of our equivalence algorithm, it is not our sole theoretical contribution. 
We have also developed:
- An on-the-fly algorithm for bisimulation, and proved its correctness using greedy bisimulation, a novel coalgebra construction.
- Two distinct methods to construct symbolic GKAT coalgebras, while also proved their correctness, finiteness, and correspondence.

# Detailed Response

### Reviewer 1

> why is it relevant to check whether two GKAT expressions are equivalent?

GKAT has been used in several important verification tasks in the domain of networks, control-flow transformation, probabilistic computation, and program logic.
We mentioned this fact at the end of the related work section. 
We will also mention the practical application of GKAT in the introduction section.

> it's unclear to me what section III is for if you have section IV

The correctness result of section IV is based on the correctness result in section III. 
Besides, there are also cases where the number of primitive tests are small and the algorithm in section III can potentially be more efficient.

### Reviewer 2

> there can still be dead states in a normalized coalgebra

Dead states are indeed preserved in the normalized coalgebra.
This approach makes the normalized coalgebra share the same carrier as the original coalgebra, allowing us to define the semantics of dead states also via the unique map into final coalgebra.

> can you say a bit more about SymKAT and why it might perform worse

SymKAT is a symbolic equivalence checker for Kleene Algebra with Tests (KAT).
Our algorithm is specialized to GKAT, thus able to leverage the deterministic structure to produce smaller automata (coalgebra) and use existing SAT solver to resolve boolean equivalence/inequivalence.
Our contribution, to the best of our knowledge, is both a theoretical and practical demonstration that GKAT equivalence can be checked more efficiently using a specialized solver.

### Reviewer 3

> Theorem 23 is true, but Theorem 23 is not the proper tool to justify the correctness of the algorithm as mentioned on p7c2

A union-find structure represents an equivalence relation. 
By employing union-find instead of a set of equivalent pairs, algorithm will search for a greedy bisimulation that is also an equivalence relation. 
While the correctness of greedy bisimulation is established (Theorem 21), there is no inherent justification for why finding a greedy bisimulation that is also an equivalence relation is equivalent to finding just any greedy bisimulation. 
To address this, we will apply Theorem 23 to demonstrate that there exists a greedy bisimulation that is also an equivalence relation if and only if two states are trace equivalent.

> p5c2 "This approach offers several benefits: first, the liveness detection is only called when necessary, thus unlikely to iterate through the entire coalgebra; second, the algorithm can short-circuit on a counter-example, finally [...]""
> -> The first two are the same benefit.

The first benefit stems from the laziness of our liveness checking algorithm, allowing us to iterate through fewer states in the classical algorithm (see overall response and the next comment).  
This benefit is apparent no matter the result or the generation method of the coalgebra. 
The second benefit stems from the on-the-fly nature of our algorithm.
Specifically, our algorithm can terminate without generating the rest of the coalgebra, when a counter-example is encountered while applying an on-the-fly generation method (such as derivatives).
In contrast, the original algorithm requires the complete generation of the coalgebra in order to perform liveness checking first.

> p7c2 "In the extreme case when the two input states are infinite-trace equivalent, the on-the-fly algorithm can even skip liveness checking entirely." 
> -> yes, but in that case the liveness analysis would be faster (linear instead of almost linear)

We are uncertain why the original algorithm might be faster in this specific case. 
The original algorithm requires determining the liveness of every single state before performing bisimulation. 
In contrast, our algorithm only conducts liveness checking on a few states when the bisimulation algorithm detects a discrepancy.
Consider two bisimilar states (not just finite-trace equivalent) in a GKAT coalgebra composed entirely of live states. 
The classical algorithm would first perform a liveness check, iterating through all the states, and then proceed with bisimulation, iterating through all the states again. 
Our algorithm, however, simply performs the bisimulation and confirms there is no discrepancy, effectively bypassing the liveness check altogether.

# !!!!!!! NOTES, DO NOT SUBMIT !!!!!!!
1. Practicality of the tool
    <!-- - Do people care about GKAT 
        - not a lot of application of GKAT, because there is no evidence that GKAT is faster than KAT -->
    - Speed up of normalization 
        - discussion at the end of section III;
        - better sketch of liveness checking.
    - Trivial/Dead Equivalence
2. Comparison with existing work
    - Intuitive, but coalgebraic characterization is novel
    - coalgebraic characterization enables straightforward proofs
    - Theorem 22 is non-trivial because there is no assumption of finiteness
3. additional results besides automata
    - We also presented two ways to generate automata in Section V and proved them correct and equivalent.
    - Automata shape difference, allow acceptation or rejection based on an atom.
3. Tests generation
    - We generate the tests roughly based on the shape of real-world program input 
        - more sequencing than while etc.
    <!-- - "trivial" tests does not impact the performance evaluation for Thompson
    - "dead" tests does not impact the performance evaluation -->
    - negative cases are random expression, but they still need to do liveness check
    - We will rework the tests to 
        - reduce the number of trivial and dead tests, 
        - make sure that the random expression do go into dead state detection immediately
4. Writing
    - More about the efficiency of laziness

### 2025-03-18 rebuttal.md

The first benefit stems from the laziness of our liveness checking algorithm, allowing us to iterate through fewer states in the classical algorithm (see overall response and the next comment).  
This benefit is apparent no matter the result or the generation method of the coalgebra. 
The second benefit stems from the on-the-fly nature of our algorithm.
Specifically, our algorithm can terminate without generating the rest of the coalgebra, when a counter-example is encountered while applying an on-the-fly generation method (such as derivatives).
In contrast, the original algorithm requires the complete generation of the coalgebra in order to perform liveness checking first.

> p7c2 "In the extreme case when the two input states are infinite-trace equivalent, the on-the-fly algorithm can even skip liveness checking entirely." 
> -> yes, but in that case the liveness analysis would be faster (linear instead of almost linear)

We are uncertain why the original algorithm might be faster in this specific case. 
The original algorithm requires determining the liveness of every single state before performing bisimulation. 
In contrast, our algorithm only conducts liveness checking on a few states when the bisimulation algorithm detects a discrepancy.
Consider two bisimilar states (not just finite-trace equivalent) in a GKAT coalgebra composed entirely of live states. 
The classical algorithm would first perform a liveness check, iterating through all the states, and then proceed with bisimulation, iterating through all the states again. 
Our algorithm, however, simply performs the bisimulation and confirms there is no discrepancy, effectively bypassing the liveness check altogether.

<!-- 
!!!!!!! NOTES, DO NOT SUBMIT !!!!!!!
1. Practicality of the tool
    - Speed up of normalization 
        - discussion at the end of section III;
        - better sketch of liveness checking.
    - Trivial/Dead Equivalence
2. Comparison with existing work
    - Intuitive, but coalgebraic characterization is novel
    - coalgebraic characterization enables straightforward proofs
    - Theorem 22 is non-trivial because there is no assumption of finiteness
3. additional results besides automata
    - We also presented two ways to generate automata in Section V and proved them correct and equivalent.
    - Automata shape difference, allow acceptation or rejection based on an atom.
3. Tests generation
    - We generate the tests roughly based on the shape of real-world program input 
        - more sequencing than while etc.
    - negative cases are random expression, but they still need to do liveness check
    - We will rework the tests to 
        - reduce the number of trivial and dead tests, 
        - make sure that the random expression do go into dead state detection immediately
4. Writing
    - More about the efficiency of laziness 
-->

### 2025-03-19 rebuttal-600words.md

# Overall Response

We express our gratitude for the valuable feedback received from reviewers. 
We will carefully address the points raised and implement the necessary changes to our paper.

First, we address concerns surrounding lazy normalization. 
The classical algorithm guarantees a full traversal of the coalgebra during liveness checking prior to bisimulation checking, our algorithm invokes bisimulation and only traverses the reachable states of the discrepancies.
Therefore, lazy liveness checking consistently visits fewer states, even for positive results. 
Additionally, lazy liveness checking enables on-the-fly coalgebra generation using derivatives.

While our algorithms seem intuitive, our proofs are not trivial. 
The simple proofs are enabled by the design of greedy bisimulation. 
Without this middle ground, verifying Theorem 22 would be challenging, especially given that neither the input coalgebras nor their bisimulation are assumed to be finite.

Finally, we emphasize that our contribution extends beyond bisimulation of coalgebras.
We also presented two methods for generating coalgebras from expressions and proved their correctness, finiteness, and correspondence.

Regarding our experiments, we have revised our tests in our repository. 
We also stress that trivial tests do not provide any performance benefit for algorithms based on Thompson's construction, and dead tests do not favor our algorithm. 
As mentioned before, we have conceptual demonstrations of performance improvements in the submitted version.

# Reviewer Questions

### Reviewer 1

> Where are the benchmarks from?

It is generated by https://github.com/c-cube/qcheck, based on algebraic rules of GKAT (excluding fixpoint rule) with reflexivity, congruence, and symmetry rules.
We will include the generation script and provide documentation.

> Unnecessary but harmful generalization

We will clarify that the functor `F` satisfies necessary properties.

### Reviewer 2

Our approach diverges from previous works on symbolic transducers in several aspects:
- Our on-the-fly method is tailored for normalization, a unique challenge in achieving finite trace equivalence for GKAT. 
- The signature of GKAT differs from traditional symbolic transducers. Unlike typical accepting or rejecting states, a GKAT state can accept or reject based on different input atoms. This results in a semantics where a trace ends in an atom, representing the output program state.

While symbolic bisimulation is a valuable contribution, it is not our sole contribution. 
We spent most of the paper developing an on-the-fly bisimulation algorithm and two distinct method to construct symbolic GKAT coalgebra from expressions, and proved their desirable properties.

# Detailed Response

### Reviewer 1

> it's unclear to me what section III is for

The correctness result of section IV is based on that of section III.

### Reviewer 2

> can you say a bit more about SymKAT

SymKAT is a symbolic equivalence checker for KAT; on the other hand, our algorithm focuses on the GKAT fragment.
This specialization allows us to produce smaller coalgebra and utilize efficient SAT solver.
Our comparison demonstrated that GKAT equivalence can be checked more efficiently using our specialized solver.

### Reviewer 3

> Theorem 23 is not the proper tool to justify the correctness of the algorithm

Union-find structures represent equivalence relations. 
By employing union-find instead of a set of pairs, our algorithm searches for a greedy bisimulation that is also an equivalence relation. 
While the correctness of greedy bisimulation is established (Theorem 21), Theorem 23 justifies finding a greedy bisimulation that is also an equivalence relation is correct.

> p7c2 [...] the liveness analysis would be faster

Consider two bisimilar states in a GKAT coalgebra with no dead state. 
The classical algorithm would perform a liveness check, iterating through all the states; then proceed with bisimulation, iterating through all the states again. 
Our algorithm simply performs the bisimulation, bypassing the liveness check altogether.

<!-- > p5c2 [...] The first two are the same benefit.

The first benefit stems from the laziness of our liveness checking algorithm, allowing us to iterate through fewer states in the classical algorithm (see overall response and the next comment).  
This benefit is apparent no matter the result or the generation method of the coalgebra. 
The second benefit stems from the on-the-fly nature of our algorithm.
Specifically, our algorithm can terminate without generating the rest of the coalgebra, when a counter-example is encountered while applying an on-the-fly generation method (such as derivatives).
In contrast, the original algorithm requires the complete generation of the coalgebra in order to perform liveness checking first. -->

<!-- 
!!!!!!! NOTES, DO NOT SUBMIT !!!!!!!
1. Practicality of the tool
    - Speed up of normalization 
        - discussion at the end of section III;
        - better sketch of liveness checking.
    - Trivial/Dead Equivalence
2. Comparison with existing work
    - Intuitive, but coalgebraic characterization is novel
    - coalgebraic characterization enables straightforward proofs
    - Theorem 22 is non-trivial because there is no assumption of finiteness
3. additional results besides automata
    - We also presented two ways to generate automata in Section V and proved them correct and equivalent.
    - Automata shape difference, allow acceptation or rejection based on an atom.
3. Tests generation
    - We generate the tests roughly based on the shape of real-world program input 
        - more sequencing than while etc.
    - negative cases are random expression, but they still need to do liveness check
    - We will rework the tests to 
        - reduce the number of trivial and dead tests, 
        - make sure that the random expression do go into dead state detection immediately
4. Writing
    - More about the efficiency of laziness 
-->

### 2025-03-19 rebuttal-600words.md

First, we address concerns surrounding lazy normalization. 
The classical algorithm guarantees a full traversal of the coalgebra during liveness checking prior to bisimulation checking, our algorithm invokes bisimulation and only traverses the reachable states of the discrepancies.
Therefore, lazy liveness checking consistently visits fewer states, even for positive results. 
Additionally, lazy liveness checking enables on-the-fly coalgebra generation using derivatives.

While our algorithms seem intuitive, our proofs are not trivial. 
The simple proofs are enabled by the design of greedy bisimulation. 
Without this middle ground, verifying Theorem 22 would be challenging, especially given that neither the input coalgebras nor their bisimulation are assumed to be finite.

Finally, we emphasize that our contribution extends beyond bisimulation of coalgebras.
We also presented two methods for generating coalgebras from expressions and proved their correctness, finiteness, and correspondence.

Regarding our experiments, we have revised our tests in our repository. 
We also emphasize that neither trivial nor dead tests provide performance benefit in our current implementation, so submitted tests are reflective of the general cases. 
As mentioned before, we have conceptual demonstrations of performance improvements in the submitted version.

### 2025-03-20 rebuttal.md

# Overall Response

We would like to express our sincere gratitude for the constructive feedback received from the reviewer. 
We will carefully address the points raised and implement the necessary changes to our paper.

First, we address the efficiency of our lazy normalization approach. 
The classical algorithm always performs a full traversal of the coalgebra during liveness checking prior to bisimulation checking, our algorithm invokes bisimulation and only run traverses the reachable states of the discrepancy state.
Therefore, our algorithm consistently traverses fewer states for liveness checking, even for positive results. 
Additionally, lazy liveness checking enables on-the-fly coalgebra generation using methods like derivatives, which the original algorithm does not support.

Second, we address the theoretical contribution of our work. 
While the result may seem intuitive, we maintain that our proof is not trivial. 
In particular, directly connecting the algorithm with bisimulation on the normalized coalgebra will be challenging (similar to Theorem 20), since we assume neither the input coalgebra nor their bisimulation is finite.
We have reworked the framework several times to pin-down the intuitive and concise coalgebra structure "greedy bisimulation" to sever as a middle ground, so that most of the proof are either unfolding definitions or results of standard coalgebra manipulations.

@@

Finally, we emphasize that our contribution extends beyond bisimulation of machines (or coalgebras).
In Section V, we present two methods for generating coalgebras from expressions and prove their correctness, finiteness, and correspondence.
The finiteness result, in particular, arises naturally from the correspondence result. 
The simplicity of this result is noted by one of the reviewers.

Regarding experiments in our work, we have updated our test cases with all the "trivial" or dead test cases removed, this development can be found in our repository: https://anonymous.4open.science/r/rust-gkat-071E/ .
Evidently, our performance are minimally impacted by the new test cases. 
We hypothesize that because our implementation is so fast, non-negligible amount of time are spent on parsing and hash-consing expressions.
Plus, trivial test cases do not provide any performance benefit for our algorithm based on Thompson's construction, and dead test cases also do not favor our algorithm against other tools.

@@

# Reviewer Questions

### Reviewer 1

> Where are the benchmarks from?

It is generated by https://github.com/c-cube/qcheck, based on congruence, symmetry, and some algebraic rules of GKAT (excluding rules that will be "trivial" tests, as named by Reviewer 3).
We will include the generation script in the final submission, and provide documentation on test generation

@@

> why is it relevant to check whether two GKAT expressions are equivalent?

GKAT has been used in several important verification tasks in the domain of networks, control-flow transformation, probabilistic computation, and program logic.
We mentioned this fact at the end of the related work section. 
We will also mention the practical application of GKAT in the introduction section.

> it's unclear to me what section III is for if you have section IV

The correctness result of section IV is based on the correctness result in section III. 
Besides, there can be edge cases (depending on the application) where the number of primitive tests are so small that the algorithm in section III can potentially be more efficient.

@@

### Reviewer 2

> there can still be dead states in a normalized coalgebra

As in all the works in GKAT, dead states are indeed preserved in the normalized coalgebra.
This approach makes the normalized coalgebra share the carrier as the original coalgebra, allowing the semantics of states to be defined via the unique map into final coalgebra, as it is standard.

@@

> can you say a bit more about SymKAT and why it might perform worse

SymKAT is a symbolic equivalence checker for Kleene Algebra with Tests (KAT).
Our algorithm is specialized to GKAT, thus able to leverage the deterministic structure to produce smaller automata (coalgebra) and use existing SAT solver to resolve boolean equivalence/inequivalence.
Our work is both the first theoretical and practical demonstration that GKAT equivalence can be checked more efficiently than KAT, by using a specialized solver like we designed in this paper.

> GKAT is a bit artificial

GKAT semantics corresponds to trace semantics, a standard semantics of programming language and control-flow analysis.
Indeed, as you specified, trace equivalence is a stronger equivalence than input-output equivalence.
However, trace equivalence turns out to be useful in various applications, as we listed in the last paragraph of the related works section, and in response to reviewer 1.

@@

### Reviewer 3

> Theorem 23 is true, but Theorem 23 is not the proper tool to justify the correctness of the algorithm as mentioned on p7c2

A union-find structure represents an equivalence relation. 
By employing union-find instead of a set of state pairs, our algorithm will effectively search for a greedy bisimulation that is also an equivalence relation. 
While the correctness of greedy bisimulation is established (Theorem 21), there is no inherent justification for why finding a greedy bisimulation that is also an equivalence relation is equivalent to finding just any greedy bisimulation. 
To address this, we will apply Theorem 23 to demonstrate that there exists a greedy bisimulation that is also an equivalence relation if and only if two states are trace equivalent.

@@

The first benefit stems from the laziness of our liveness checking algorithm, allowing us to iterate through fewer states in the classical algorithm (see overall response and the next comment).  
This benefit is apparent no matter the result or the generation method of the coalgebra. 
The second benefit stems from the on-the-fly nature of our algorithm.
Specifically, our algorithm can terminate without generating the rest of the coalgebra, when a counter-example is encountered while applying an on-the-fly generation method (such as derivatives).
In contrast, the original algorithm requires the complete generation of the coalgebra in order to perform liveness checking first.

> p7c2 "In the extreme case when the two input states are infinite-trace equivalent, the on-the-fly algorithm can even skip liveness checking entirely." 
> -> yes, but in that case the liveness analysis would be faster (linear instead of almost linear)

We are uncertain why the original algorithm might be faster in this specific case. 
The original algorithm requires determining the liveness of every single state before performing bisimulation. 
In contrast, our algorithm only conducts liveness checking on a few states when the bisimulation algorithm detects a discrepancy.
Consider two bisimilar states (not just finite-trace equivalent) in a GKAT coalgebra composed entirely of live states. 
The classical algorithm would first perform a liveness check, iterating through all the states, and then proceed with bisimulation, iterating through all the states again. 
Our algorithm, however, simply performs the bisimulation and confirms there is no discrepancy, effectively bypassing the liveness check altogether.

> say somewhere that these are more Antimorov' partial derivatives than Brzozowski's derivatives

Although we do have power set in the signature of symbolic GKAT coalgebra, our derivative do not transition in a non-deterministic manner.
In fact, we are able to out-perform state-of-the-art tools for KAT equivalence precisely because the GKAT fragment is deterministic.
We believe that our derivative is closer to Brozozowski style derivative, as we will transition to (at most) a single expression given an input expression, as opposed to a set of expression.

### 2025-03-20 rebuttal.md

# Overall Response

We would like to express our sincere gratitude for the constructive feedback received from the reviewer. 
We will carefully address the points raised and implement the necessary changes to our paper.

First, we address the efficiency of our lazy normalization approach. 
The classical algorithm performs a full traversal of the coalgebra during liveness checking before bisimulation checking. 
In contrast, our algorithm invokes bisimulation and only traverses the reachable states of the discrepancy. 
This results in fewer states being traversed during liveness checking, even for positive results. 
Additionally, lazy liveness checking enables on-the-fly coalgebra generation using methods like derivatives, which the original algorithm does not support.

Second, we address the theoretical contribution of our work. 
While the result may seem intuitive, our proof is not trivial. 
Directly connecting the algorithm with bisimulation on the normalized coalgebra, similar to Theorem 21, is challenging; partially because we do not assume that the input coalgebra or bisimulation are finite.
Our development leads to a more general result than the finitary case. 
We have reworked the framework several times to pin-down the intuitive notion of "greedy bisimulation", leading to modular results that build up to the final correctness result. 
The effectiveness of greedy bisimulation as a theoretical concept is demonstrated by the simplicity of most proofs, which are either unfoldings of definitions or standard exercises in coalgebra.

Finally, we emphasize that our contribution extends beyond bisimulation of coalgebras (or machines).
In Section V, we present two methods for generating coalgebras from expressions and prove their correctness, finiteness, and correspondence. 
The finiteness result elegantly arises from the correspondence result, a point noted by one of the reviewers.

Regarding experiments in our work, We have reworked our test cases, removing all "trivial" or dead test cases. 
Evidently, our performance is minimally impacted by the new test cases. 
We hypothesize that our implementation is so fast that a non-negligible amount of time is spent on parsing and hash-consing expressions, causing the performance impact of "trivial" tests to be minimal. 
Additionally, "trivial" tests do not provide any performance benefit for the algorithm based on Thompson's construction, and dead test cases do not favor our algorithms against other tools.

@@

The correctness result of section IV is based on the correctness result in section III. 
Besides, there can be edge cases (depending on the application) where the number of primitive tests are so small that the algorithm in section III can potentially be more efficient.

### Reviewer 2

> there can still be dead states in a normalized coalgebra

As in all the works in GKAT, dead states are indeed preserved in the normalized coalgebra.
This approach allows the semantics of dead states to be defined in the standard manner, i.e. via the unique map into final coalgebra.

@@

> can you say a bit more about SymKAT and why it might perform worse

SymKAT is a symbolic equivalence checker for Kleene Algebra with Tests (KAT). 
Our algorithm is specialized for GKAT, leveraging its deterministic structure to produce smaller automata (coalgebras) and utilizing existing SAT solvers to resolve boolean equivalence and inequivalence. 
This work represents the first theoretical and practical demonstration that GKAT equivalence can be checked more efficiently than KAT by employing a specialized solver, as designed in this paper.

@@

> GKAT is a bit artificial

GKAT semantics is based on trace semantics, a standard semantics used in programming languages and control-flow analysis. 
As you noted, trace equivalence is a stronger equivalence than input-output equivalence. 
However, trace equivalence proves to be useful in various applications, as we detailed in the last paragraph of the related work section and in response to Reviewer 1.
In fact, we foresee that our works here can be adapted to these applications and greatly improve their performance.

@@

> Theorem 23 is true, but Theorem 23 is not the proper tool to justify the correctness of the algorithm as mentioned on p7c2

A union-find structure represents an equivalence relation. 
By employing union-find instead of a set of state pairs, our algorithm will effectively search for a greedy bisimulation that is also an equivalence relation. 
While the correctness of greedy bisimulation is established (Theorem 21), there is no inherent justification for why finding a greedy bisimulation that is also an equivalence relation is equivalent to finding just any greedy bisimulation. 
To address this, we will apply Theorem 23 to demonstrate that there exists a greedy bisimulation that is also an equivalence relation if and only if two states are trace equivalent.

> p5c2 "This approach offers several benefits: first, the liveness detection is only called when necessary, thus unlikely to iterate through the entire coalgebra; second, the algorithm can short-circuit on a counter-example, finally [...]""
> -> The first two are the same benefit.

The first benefit stems from the laziness of our liveness checking algorithm, allowing us to iterate through fewer states in the classical algorithm (see overall response and the next comment).  
This benefit exists regardless of the result of the algorithm or the generation method of the coalgebra. 
The second benefit stems from the on-the-fly nature of our algorithm.
Specifically, while we encounter a counter-example while using derivative to generate the coalgebra on-the-fly, our algorithm can terminate without generating the rest of the coalgebra.
In contrast, the original algorithm requires the complete generation of the coalgebra in order to perform liveness checking first.

@@

> p7c2 "In the extreme case when the two input states are infinite-trace equivalent, the on-the-fly algorithm can even skip liveness checking entirely." 
> -> yes, but in that case the liveness analysis would be faster (linear instead of almost linear)

We are uncertain why the original algorithm might be faster in this specific case. 
Consider two bisimilar states (infinite-trace equivalent) in a GKAT coalgebra with no dead state. 
The classical algorithm would perform a liveness check, iterating through all the states; then proceed with bisimulation, iterating through all the states again. 
Our algorithm simply performs the bisimulation; because these two states are bisimilar, our algorithm will bypass the liveness check altogether.

@@

> say somewhere that these are more Antimorov' partial derivatives than Brzozowski's derivatives

While we employ the power set in the signature of the symbolic GKAT coalgebra, our derivatives do not transition in a non-deterministic manner. 
Specifically, when given an GKAT expression and a (valid) boolean expression, derivative will deterministically transition to a GKAT expression.
In this sense, our notion of derivative is closer to Brzozowski style than that of Antimorov.

### 2025-03-20 main.tex
minor bug
typo fix

\begin{abstract}
We present several new algorithms to efficiently decide Guarded Kleene Algebra with Tests (GKAT) equivalences.
Although the current algorithm scales nearly-linearly with the size of the expression, its performance is hindered by the need to compute the entire coalgebra for the sake of normalization and exponential performance degradation when expanding the set of primitive tests.
To address these problems, we introduce two new algorithms for bisimulation checking: an on-the-fly bisimulation algorithm based on greedy bisimulation, which enables immediate termination upon discovering a counter-example and supports on-the-fly coalgebra generation with derivatives; and a symbolic algorithm that builds upon the on-the-fly algorithm while leveraging efficient SAT solvers to speed up boolean comparisons and compress the transition structures of GKAT coalgebra.
Additionally, to utilize the symbolic bisimulation checking, we provide two methods to generate symbolic GKAT coalgebras from GKAT expressions, analogous to derivative and Thompson's construction in previous works.
We have proven the correctness of our algorithms and conducted brief complexity analyses.
Finally, we implemented the symbolic algorithms in Rust, demonstrating orders-of-magnitude performance improvements over existing tools.
\end{abstract}

@@

Given any bisimulation \({∼} ⊆ S × U\), we can construct a GKAT coalgebra \(∼_d\) s.t. it is \emph{almost} a greedy bisimulation, i.e. it is closed under \(δ_≈\) in~\Cref{def:greedy-bisim}, and contains \(∼\), except it is not necessarily the least GKAT coalgebra that satisfies these properties.

@@

We name \(ϵ̂\) the accepting function because intuitively a state \(s\) accepts an atom \(α\) when there exists a \(b ∈ ϵ̂(s)\), s.t. \(α ≤ b\); similarly, \(δ̂\) is called the transition function because a state \(s\) transitions to \(s'\) via atom \(α\) while executing \(p\) when there exists \((b, s', p) ∈ δ̂(s)\) and \(α ≤ b\).
With the above intuition in mind, a symbolic GKAT coalgebra \(Ŝ ≜ (S, ϵ̂, δ̂)\) can be lowered into a GKAT coalgebra \(S ≜ (S, δ)\) in the following manner:
\begin{align}\label{cons:lowering}
δ(s, α) ≜ \begin{cases}
    \accept & ∃ b ∈ ϵ̂(s), α ≤ b \\[5px]
    (s', p) & 
        \begin{aligned}
            & ∃ b ∈ \BExp_T, α ≤ b \\
            & \text{ and } δ(s, b) = (s', p)  
        \end{aligned}\\[5px]
    \reject & \text{otherwise}
\end{cases}
\end{align}
The function \(δ\) is well-defined, i.e. exactly one clause can be satisfied for any \(s ∈ S\) and \(α ∈ \At\), because of the disjointedness condition.
We usually use \(S\) to denote the lowering of \(Ŝ\); and the semantics of a state \(s ∈ Ŝ\) is defined as its semantics in the lowering \(⟦s⟧_{Ŝ} ≜ ⟦s⟧_{S}.\)

>>>>

\begin{abstract}
We present several new algorithms to efficiently decide Guarded Kleene Algebra with Tests (GKAT) equivalences.
Although the current algorithm scales nearly-linearly with the size of the expression, its performance is hindered by the need to compute the entire coalgebra for the sake of normalization and exponential performance degradation when expanding the set of primitive tests.
To address these problems, we introduce two new algorithms for bisimulation checking: an on-the-fly bisimulation algorithm based on greedy bisimulation, a coalgebriac representation of our algorithm; and its symbolic version. 
Our algorithm enables immediate termination upon discovering a counter-example and supports on-the-fly coalgebra generation with derivatives; and a symbolic algorithm that builds upon the on-the-fly algorithm while leveraging efficient SAT solvers to speed up boolean comparisons and compress the transition structures of GKAT coalgebra.
Additionally, to utilize the symbolic bisimulation checking, we provide two methods to generate symbolic GKAT coalgebras from GKAT expressions, analogous to derivative and Thompson's construction in previous works.
We have proven the correctness of our algorithms and conducted brief complexity analyses.
Finally, we implemented the symbolic algorithms in Rust, demonstrating orders-of-magnitude performance improvements over existing tools.
\end{abstract}

@@

Given any bisimulation \({∼} ⊆ \norm(S) × \norm(U)\), we can construct a GKAT coalgebra \(∼_d\) s.t. it is \emph{almost} a greedy bisimulation, i.e. it is closed under \(δ_≈\) in~\Cref{def:greedy-bisim}, and contains \(∼\), except it is not necessarily the least GKAT coalgebra that satisfies these properties.

@@

We name \(ϵ̂\) the accepting function because intuitively a state \(s\) accepts an atom \(α\) when there exists a \(b ∈ ϵ̂(s)\), s.t. \(α ≤ b\); similarly, \(δ̂\) is called the transition function because a state \(s\) transitions to \(s'\) via atom \(α\) while executing \(p\) when there exists \((b, s', p) ∈ δ̂(s)\) and \(α ≤ b\).
With the above intuition in mind, a symbolic GKAT coalgebra \(Ŝ ≜ (S, ϵ̂, δ̂)\) can be lowered into a GKAT coalgebra \(S ≜ (S, δ)\) in the following manner:
\begin{align}\label{cons:lowering}
δ(s, α) ≜ \begin{cases}
    \accept & ∃ b ∈ ϵ̂(s), α ≤ b \\[5px]
    (s', p) & 
        \begin{aligned}
            & ∃ b ∈ \BExp_T, α ≤ b \\
            & \text{ and } δ̂(s, b) = (s', p)  
        \end{aligned}\\[5px]
    \reject & \text{otherwise}
\end{cases}
\end{align}
The function \(δ\) is well-defined, i.e. exactly one clause can be satisfied for any \(s ∈ S\) and \(α ∈ \At\), because of the disjointedness condition.
We usually use \(S\) to denote the lowering of \(Ŝ\); and the semantics of a state \(s ∈ Ŝ\) is defined as its semantics in the lowering \(⟦s⟧_{Ŝ} ≜ ⟦s⟧_{S}.\)
